RVD2: An ultra-sensitive variant detection model for heterogeneous population sequencing

### 2013-09-11 doc/bioinformatics2013/RVD2.tex
Finished result section on ROC curves. Update Notebook on ROC experiment.

The null hypothesis is tested using power-divergence family of statistics proposed by Cressie and Read (1984). Considering observed frequencies $\{r_i; i=1,...,k\}$ and expected frequencies $\{E_i; i=1,...,k\}$, by definition the power-divergence family of statistcs indexed by real parameter $\lambda$ is 
\begin{equation}
 2nI^\lambda = \frac{2}{\lambda(\lambda+1)}\sum_{i=1}^k r_i \left[\left(\frac{r_i}{E_i}\right)^\lambda-1\right];\lambda \in R
\end{equation}

Person's $X^2 (\lambda = 1)$ statistics, the log likelih ratio statistics $(\lambda = 0)$, the Freeman-Tukey statistics $(\lambda = -1/2)$ the Neyman modified $X^2 (\lambda = -2)$ are special cases of the family. Cressie and Read (1984) also proposed that when the $\lambda =2/3$, the new power-divergence statistics outcompete the traditional power-divergence statistics.

@@

Finally, if read counts in one position passes the Bayesian Hypothesis Testing and fails the Goodness-of-fit Testing, the position will be called.

\subsubsection{Posterior distribution difference sampling}\

@@

picard to thin segment data, showing performance as read depth decreasing

\begin{figure}[h]
\begin{center}
%\includegraphics[width=160mm]{eps_figs/ROC_without_chi2.eps}
\caption{ROC curve varying read depth showing detection performance of model with Bayesian Hypothesis Test.}
\label{fig:ROC_without_chi2}
\end{center}
\end{figure}

@@

We tested the mutation detection power of our graphical model on synthetic dataset with dilution rates $0.1\%$, $0.3\%$, $1.0\%$, $10.0\%$ and read depth varying from $10^2$ to $10^5$.Different read depth dataset was achieve by thining dataset with read depth up to $10^6$ using Picard.The model had high detection power when the read depth was at $10^5$ level. At dilution rate at $0.1\%$, the model achieved performance with sensitivity at 99.74\% and specificity at 92.86\%. We missed one mutation at position 205 and had one false positive at position 8. The threshold was chosen at 0.022\% which achieved the optimal sensitivity and specificity in the ROC curve. As expected, the detection power of the graphical model decreased when the read depth was reduced.

\begin{figure}[h] %needs to go to supplementary document.
\begin{center}
\includegraphics[width=160mm]{pdf_figs/ROC_subplots_with_chi2.pdf}
\caption{ROC curve varying read depth showing method performance with Bayesian baHypothesis Testing and Chi-Square Test}
\label{fig:R}
\end{center}
\end{figure}
We also evaluateed the performance of the graphical model with Chi-square test included in the hypothesis testing.

Low read depth limited the False positive rate. we are confident in the callings

High read depth Not much influence. do not bring about bad influence

>>>>

The null hypothesis is tested using power-divergence family of statistics proposed by Cressie and Read (1984).By definition, the power-divergence family of statistcs parameterized with $\lambda$ is 
\begin{equation}
 2nI^\lambda = \frac{2}{\lambda(\lambda+1)}\sum_{i=1}^k r_i \left[\left(\frac{r_i}{E_i}\right)^\lambda-1\right];\lambda \in R,
\end{equation}
where $\{r_i; i=1,...,k\}$ are the observed frequencies and $\{E_i; i=1,...,k\}$ are the  expected frequencies.

Person's $\chi^2 (\lambda = 1)$ statistics, the log likelih ratio statistics $(\lambda = 0)$, the Freeman-Tukey statistics $(\lambda = -1/2)$ the Neyman modified $X^2 (\lambda = -2)$ are special cases of the family. Cressie and Read (1984) also proposed that when the $\lambda =2/3$, the new power-divergence statistics outcompete the traditional power-divergence statistics.

@@

\subsection{Posterior distribution difference sampling}\

@@

@@

We tested the mutation detection power of our graphical model on synthetic dataset using a two-stage test scheme with dilution rates $0.1\%$, $0.3\%$, $1.0\%$, $10.0\%$ and read depth varying from $10^2$ to $10^5$.Different read depth dataset was achieve by thining dataset with read depth up to $10^6$ using Picard. $\chi^2$ test, the second stage test can be optionally included based on average read depth.

Excluding $\chi^2$ test, the ROC curves of our graphical model using Bayesian Hypothesis Testing with different threshold $\tau$ is shown in Figure??. The model had high detection power when the read depth was at $10^5$ level. At dilution rate at $0.1\%$, the model achieved performance with sensitivity at 99.74\% and specificity at 92.86\%. We missed one mutation at position 205 and had one false positive at position 8. The threshold was chosen at 0.022\% which achieved the optimal sensitivity and specificity in the ROC curve. As expected, the detection power of the graphical model decreased when the read depth was reduced.

\begin{figure}[h] %needs to go to supplementary document.
\begin{center}
\includegraphics[width=160mm]{pdf_figs/ROC_subplots_with_chi2.pdf}
\caption{ROC curve varying read depth showing method performance with Bayesian Hypothesis Testing and Chi-Square Testing}
\label{fig:R}
\end{center}
\end{figure}

Figure ?? shows the ROC curves when both Bayesian Hypothesis Testing and $\chi^2$ testing are carried out for variant calling. Comparing with Figure ??, it can be seen that when read depth is low (from $10$ to $10^3$), the false positive rate is severely limited to zero, which means we are very confident in our variant callings. However, the low false positve rate can be achieved at a sacrifice of low true postive rate especially when the dilution rate and read depth are low. The $\chi^2$ test is a stringent test when the read depth is low.

However, if the read depth is high enough (up to $10^4$ or higher), the $\chi^2$ test can limit the false positive rate in a slight degree but no longer at a sacrifice of True Positive Rate. Actually the performance is not much influenced by $\chi_2$ testing with high read depth.
